---
title: "Building a Dynamic TOPMODEL"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building a Dynamic TOPMODEL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
                    collapse = TRUE,
                    comment = "#>"
                  )

```

The purpose of this vignette is to provide an outline of the steps needed to
build a dynamic TOPMODEL implementation using the dynatopGIS package.

# Implimentation notes

The dynatopGIS package impliments a structured, object orientated, data
flow. The steps outlined below create a `dynatopGIS` catchment object to which
actions are then applied to generate a model. 

The `dynatopGIS` package is written using the object orientated framework
provided by the `R6` package. This means that some aspects of working with the
objects may appear ìdiosymatic for some R users. In using the package as
outliend in this vignette these problems are largely obscured, except for the
call structure. However before adapting the code, or doing more compelx analysis
users should read about `R6` class objects (e.g. in the `R6` package vignettes
or in the Advanced R book). One particuler gotya is when coping an object. Using

```{r eval=FALSE}
my_new_object  <-  my_object
```

creates a pointer, that is altering `my_new_object` also alters
`my_object`. To create a new independent copy of `my_object` use

```{r eval=FALSE}
my_new_object  <-  my_object$clone()
```

# Getting started

The dynatopGIS packages works through a number of steps to generate a dynamic
TOPMODEL object suitabel for use in with the `dynatop` package. Each steps
generates one or more layers which are saved as raster or shape files into the projects
working directory (which is not nessecarily the R working directory). A record
of these layers is kept in the json format meta data file.

This vignette demonstrates the use of the `dynatopGIS` package using data from the Swindale
catchment in the UK. This is included in the `dynatopData` package.

To start first load the library

```{r load_library}
library("dynatopGIS")
```

For this vignette we will store the data into a tempory dirctory

```{r tempory_dir}
demo_dir <- tempfile("dygis")
dir.create(demo_dir)
```
and initialise the analysis by creating a new object specifying the loaction
of the meta data file, which will be created if it doesn't exist.

```{r, initialization}
ctch <- dynatopGIS$new(file.path(demo_dir,"meta.json"))
```

# Adding catchment data

The basis of the analysis are a rasterised Digital Elevation Model (DEM) of
the catchment and a vectorised representation of the river network with
attributes. Current these can be in any format supported by the ```raster```
and ```sp``` libraries respectivly.

However within the calculations used for sink filling,
flow routing and topographic index calculations the raster DEM is presumed
to be projected so that is has square cells such that
the difference between the cell centers (in meters) does not alter.

For Swindale the suitable DEM and channel files can be found using:

```{r, initialization}
dem_file <- system.file("extdata", "SwindaleDTM4mFilled.tif", package="dynatopData", mustWork = TRUE)
channel_file <- system.file("extdata", "SwindaleRiverNetwork.shp", package="dynatopData", mustWork = TRUE)
```

Either the DEM or channel files can be added to the project first. In this
case we add the DEM with

```{r, create_brick}
ctch$add_dem(dem_file)
```
*Note:* An additional row and column of `NA` is added to each edge of the
DEM. All raster layers in the project are of the same projection,
resolution and extent of the extended DEM.

Adding river channel data is more complex. The `add_channel` method requires a
`SpatialLinesDataFrame` or `SpatialPolygonsDataFrame` as generated by the `sp`
package. Each entry in the data.frame is treated as a length of river channel
which requires the following properties

-   *endNode* - a label for the downstream end of the river length
-   *startNode* - a label for the upstream end of the river length
-   *length* - the length in meters

Additional proporties are currently kept but ignored with two exceptions:
-   *id* - this is copied to `original_id` with a warning since `id` is used internally
-   *width* - if the channel is specified with a line sections then the *width*
    property is used to buffer the lines to create channle polygons.
	
Since it is possible that these properties are present in a data file
under different names the `add_channel` method allows for
renaming. To illustrate this let us examine the river network for Swindale

```{r, channel_current}
sp_lines <- rgdal::readOGR(channel_file)
head(sp_lines)
```

The main properties are present under appropriate names and a call to
the `add_channel` method with no options would be successful. However if we want to carry over the addition information in the "indentifier" as "channel_id" a
named vector giving the variable names to be used could be provided. In this
case:

```{r, channel_properties}
property_names <- c(channel_id="identifier",
                    endNode="endNode",
                    startNode="startNode",
                    length="length")
```

The river network can
then be created in the correct format by

```{r, add_channel}
ctch$add_channel(sp_lines,property_names)
```

Since the data set for Swindale does not contain a channel width the default
width of 2m is used.

# Computing Basic Properties

So far the DEM and channel data exist in isolation. Next we compute some basic
summaries for each square in the DEM, specifically:
-   land area - the area in the DEM cell covered by land
-   channel area - the area in the DEM cell covered by channel
-   channel id - the id of the channel within the cell, corresponding to
the id in the channel object.

If multiple river lengths intersect a DEM cell the properties of the channel
length with the largest area of intersection are used.

This computation is done by calling

```{r basic_properties}
ctch$compute_area()
```

# Getting and plotting catchment information

The `dynatopGIS` class has methods for returning and plotting the GIS data in
the project. The names of all the different GIS layers stored is returned by

```{r, list_layers}
ctch$get_layer()
```

These can be plotted (with or without the channel), for example

```{r, plot}
ctch$plot_layer("dem", add_channel=TRUE)
```

or returned, for example

```{r, get_layer}
tmp <- ctch$get_layer("dem")
```
All layers are retuened as `RasterLayer` objects with the exception of the
`channel` layer which is returned as a `SpatialPolygonsDataFrame` object. The
complete meta data file can be retrieved with

```{r, get_layer}
tmp <- ctch$get_meta("dem")
```


# Filling sinks

For the hill slope to be connected to the river network all DEM cells must
drain to those that intersect with the river network.

The algorithm of <TODO add ref> implimented in the `sink_fill` method ensures
this is the case. Since the algorithm is iterative the execution time of the
function is limited by capping setting a maximum number of iterations. If
this limit is reached without completion the method can call again with the
"hot start" option to continue from where it finished.

For Swindale, where the example DEM is already partially filled the algorithm
only alters a small area near the foot of the catchment.

```{r, sink_fill}
ctch$sink_fill()

raster::plot( ctch$get_layer('filled_dem') - ctch$get_layer('dem'),
             main="Changes to height")
```

# Computing properties

Two sets of properties are required for Dynamic TOPMODEL. The fist set is
those required within the evaluation of the model; gradient and contour
length. the second set are those used for dividing the catchment up into
Hydrological Response Units (HSUs). Traditionally the summary used for the
separation of the HRUs is the topographic index, which is the natural
logarithm of the upslope area divided by gradient.

These are computed using the formulea in Quinn 1991 (TO DO add ref).

The upstream area is computed by routing down slope with the fraction of the
area being routed to the next downstream pixel being proportional to the
gradient times the contour length. 

The local value of the gradient is computed using the average of a
subset of between pixel gradients. For a normal 'hill slope' cell these are the
gradients to downslope pixels weighted by contour length. In the case of
pixels which contain river channels the average of the gradients from upslope
pixels weighted by contour length us used.

These properties are computed in an algorithm that passes over
the data once in descending height. It is called as follows

```{r, calc_atb}
ctch$compute_properties()
```

The plot of the topographic index shows a pattern of increasing values closer
to the river channels

```{r, plot_atb}
## plot of topographic index (log(a/tan b))
ctch$plot_layer('atanb',add_chennal=TRUE)
```

## Adding additional layer

Properties may come in addional GIS layers. To demonstrate the addition of an
additional layer we will extract the filled dem

```{r ,extract_filled}
tmp <- ctch$get_layer("filled_dem")
```

then seperate it into a layers representing land above and below 500m.

```{r height layer}
tmp <- raster::reclassify( tmp,
                          matrix(c(0,500,NA,
                                   500,1000,-999),
                                 byrow=TRUE))
```

The resulting raster object can know be written to a file

```{r, write_height_layer}
raster::writeRaster(tmp,file.path(demo_dir,"greater_500.tif"))
```
which is added to the meta data with
```{r, add_height_layer}
ctch$add_layer("greater_500",file.path(demo_dir,"greater_500.tif"))
ctch$add_layer()
```

# Flow distances and ordering

Since `dynatop` simulations make use of ordered HRUs to work downslope, a
metric is required to order the downslope
sequencing. The calculation of four such metrics is supported

- *shortest flow length* - the shortest length based on the pixel flow paths
to a channel
- *Dominant flow length* - the distance to a channel moving in the dominant
(largest fraction) flow direction from any grid cell
- *Expected flow length* - the distance to the channel based on a weighted
average of the downlope flow lengths. Weights are given by the fraction of
flow in each direction.
- *Band* - A strict computational order, starting at the channel arranged such
that if all pixels in band $i+1,i+2,\ldots$ are evaluated the inflows to band
$i$ are known.

The computation is initiated with
```{r, flow_length}
ctch$compute_flow_length()
```

The additional layers can be examined as expected
```{r, flow_length_plot}
ctch$get_layer()
ctch$plot_layer("band")
```

# Classifying into Hydrological Similarity Units

Methods are provided for the classification of the catchment into Hydrological
 Similarity Units (HSUs). The classifications generated in this process are
 augmented with a further distance based seperation when generating a
 `dynatop` model (see following section).
 
By definition each channel length is treated as a single HSU. Two ways are
provided for dividing the hillslope up into HSUs.
The first way is *splitting* where one or more landscape properties are
divided up into classes and HSUs defined by each unique combination of
classes.
The second *burning* enforces classes onto distinct areas.

The `classify` method of a `dynatopGIS` allows the use of both methods, first
*splitting* then *burning* to create hillslope HSUs.

## Splitting

To split a catchment into HRUs the breaks between the classes need to be
specified. This is done by forming a named ```list``` of cuts. The names
correspond to layers of GIS data within the `dynatopGIS` object. The values of
the variables should be numeric and are used as follows:

- `NA`: use the layer 'as is'; i.e. presume it is already classified
- Single value: defines the number of splits which are automatically selected
- Vector of values: treated as the breaks between classes

To demonstate this HSUs for Swindale based on dividing the topographic index
into 21 classes can be gernated by

```{r, atb_split}
ctch$classify("atb_split",cuts=list(atanb=20))
```

or if the land greater then 500 in altitude is treated as a single HSU using a 'burn'

```{r, atb_split}
ctch$classify("atb_500_split",cuts=list(atanb=20),burns="greater_500")
```
Note that by giving the area to be burnt in negative value when it was
generated above we have ensured that the values do not clash whith those generated by the cuts which
(except potentially when a cut is NA) will always be positive.

The final classificationcan be plotted as with any other layer

```{r plot_class}
ctch$plot_layer("atb_500_split")
```

In the meta data the cuts and burns used to generate the layer are
recorded. these can be retreved with

```{r see_class}
ctch$get_class_method("atb_500_split")
```

# Generating a dynamic TOPMODEL

A Dynamic TOPMODEL sutiable for use with the ```dynatop``` package can be
generated using the `create_model` method. This used the current
classification to generate the model. The required model structure is
given in the vignettes of `dynatop` package and is not described here in
details. 

Since `dynatop` simulations make use of ordered HSUs to work downslope, a
seperation by a distance layer (see earlier section) which represents the ordered downslope
sequencing of the pixels is used. For the distance layer `create_model`
method takes an input in the same format as the specification of a `split`
in the `classify` method. Further can be found in the seperate vignette on
distance calculations (TODO). *It is strongly recommended that the 'band'
distance metric is used directly as shown below*.

For example in the case of the division of Swindale by topographic index into
21 classes and the bands directly the resulting model can be generated by

```{r, model_atb_split}
ctch$create_model("new_model","atb_split","band",NA)
```

Looking at the fiels within the `demo_dir` folder

```{r, model files}
list.files(demo_dir,pattern="new_model*")
```

shows that an addition raster map of the HSUs has been created in
`new_model.tif` along with a file `new_model.rds` containing a model
suitable for `dynatop`

The map of HSUs can be plotted as with any layer

```{r plot_model}
ctch$plot_layer("new_model")
```

The values on the map correspond to the ìd column of the hillslope tabel in
the `dynatop` model.
