---
title: "Building a Dynamic TOPMODEL"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building a Dynamic TOPMODEL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
                    collapse = TRUE,
                    comment = "#>"
                  )

```

The purpose of this vignette is to provide an outline of the steps needed to
build a dynamic TOPMODEL implementation using the dynatopGIS package.

# Implimentation notes

The dynatopGIS package impliments a structured, object orientated, data
flow. The steps outliend below create a `dynatopGIS` catchment object to which
actions are then applied to generate a model. At any point in the analysis the
object can be written to disk using the usual R commands such as `saveRDS`. 

The `dynatopGIS` package is written using the object orientated framework
provided by the `R6` package. This means that some aspects of working with the
objects may appear ìdiosymatic for some R users. In using the package as
outliend in this vignette these problems are largely obscured, except for the
call structure. However before adapting the code, or doing more compelx analysis
users should read about `R6` class objects (e.g. in the `R6` package vignettes
or in the Advanced R book). One particuler gotya is when coping an object. Using

```{r eval=FALSE}
my_new_object  <-  my_object
```

creates a pointer, that is altering `my_new_object` also alters
`my_object`. To create a new independent copy of `my_object` use

```{r eval=FALSE}
my_new_object  <-  my_object$clone()
```

# Getting started

First load the library

```{r load_library}
library("dynatopGIS")
```

The basis of the analysis are a rasterised Digital Elevation Model (DEM) of
the catchment and a vectorised representation of the river network with
attributes. Current these can be in any format supported by the ```raster```
and ```sp``` libraries respectivly.

However within the calculations used for sink filling,
flow routing and topographic index calculations the raster DEM is presumed
to be projected so that is has square cells such that
the difference between the cells centers (in meters) does not alter.

## Swindale

The use of the dynatopGIS package is demonstrated using data from the Swindale
catchment in the UK. This is included in the external data directory of the
package and can be accessed at the follwoing paths.
```{r, initialization}
dem_file <- system.file("extdata", "SwindaleDTM4mFilled.tif", package="dynatopGIS")
channel_file <- system.file("extdata", "SwindaleRiverNetwork.shp", package="dynatopGIS")
```

# Initialising the `dynatopGIS` objects

A `dynatopGIS` object is initialised using a RasterLayer containing the
DEM. All information about spatial extent and project are taken from this. For
the example we intialise as follows

```{r, create_brick}
dem <- raster::raster(dem_file)
ctch <- dynatopGIS$new(dem)
```

Adding river channel data is mroe complex. The `add_channel` method requires a
`SpatialLinesDataFrame` or `SpatialPolygonsDataFrame` as generated by the `sp`
package. Each entry in the data.frame is treated as a length of river channel
which requires the following properties

-   *endNode* - a label for the downstream end of the river length
-   *startNode* - a label for the upstream end of the river length
-   *length* - the length in meters

Additional proporties are currently ignored with two exceptions:
-   *id* - this is overwrittem with a warning and used internally
-   *width* - if the channel is specified with a line sections then the *width*
    property is used to buffer the lines to create channle polygons.
	
Since it is possible that these properties are present in a data file
under different names the `add_channel` method allows for
renaming. To illustrate this let us examine the river network for Swindale

```{r, channel_current}
sp_lines <- rgdal::readOGR(channel_file)
head(sp_lines)
```

The main properties are present under appropriate names and a call to
the `add_channel` method with no options would be successful. However if we want to carry over the addition information in the "indentifier" as "channel_id" a
named vector giving the variable names to be used could be provided. In this
case:

```{r, channel_properties}
property_names <- c(channel_id="identifier",
                    endNode="endNode",
                    startNode="startNode",
                    length="length")
```

Since the data set for Swindale does not contain a channel width we
use the default width of 2m in `create_channel`. The river network can
then be created in the correct format by

```{r, add_channel}
ctch$add_channel(sp_lines,property_names)
```

The ```add_channel``` method intersects
the DEM and river network to determine which of the raster cells contain
parts of the river network. From this three layers of the brick are populated:
-   land_area - the area in the DEM cell covered by land
-   channel_area - the area in the DEM cell covered by channel
-   channel_id - the id of the channel within the cell, corresponding to
the id in the channel object.

If multiple river lengths intersect a DEM cell the properties of the channel
length with the largest area of intersection are used. 

### Getting and plotting catchment information

The `dynatopGIS` class has methods for returning and plotting the GIS data.
A list of all the different GIS information stored is returned by

```{r, list_layers}
ctch$get_layer()
```

These can be plotted (with or without the channel), for example

```{r, plot}
ctch$plot_layer("dem")
```

and returned as `RasterLayer` objects

```{r, get_layer}
ctch$get_layer("dem")
```

# Filling sinks

For the hill slope to be connected to the river network the hill slope parts of
the DEM must drain to the DEM cells which intersect with the river network.

The algorithm of <TODO add ref> implimented in the `sink_fill` method ensures
this is the case. Since the algorithm is iterative the execution time of the
function is limited by capping setting a maximum number fo iterations. If
this limit is reached without completion the method can call again with the
"hot start" option to continue from where it finished.

For Swindale, where the example DEM is already partially filled the algorithm
only alters a small area near the foot of the catchment.

```{r, sink_fill}
ctch$sink_fill()

raster::plot( ctch$get_layer('filled_dem') - ctch$get_layer('dem'),
             main="Changes to height")
```

## Computing summary properties

The basis of separation of the hydrological response units in TOPMODEL is the
topographic index, which is computed from the upstream area and gradient.

These are computed using the formulea in Quinn 1991 (TO DO add ref).

The upstream area is computed by routing down slope with the fraction of the
area being routed to the next downstream pixel being proportional to the
gradient times the contour length. 

The local value of the gradient is computed using the average of a
subset of between pixel gradients. For a normal 'hill slope' cell these are the
gradients to downslope pixels weighted by contour length. In the case of
pixels which contain river channels the average of the gradients from upslope
pixels weighted by contour length us used.

These properties (and others) are computed in an algorithm that passes over
the data twice. First to compute identify 'peaks' in the dem, then to
travel downslope. It is called as follows

```{r, calc_atb}
ctch$compute_properties()

## plot of topographic index (log(a/tan b))
ctch$plot_layer('atanb')

## plot of order of pixels (1 highest)
ctch$plot_layer('band')
```

## Classifying into Hydrological Similarity Units

Methods are provided for the classification of the catchment into Hydrological
 Similarity Units (HSUs). A `dynatopGIS` object can hold an single classification at
 a time. 
 
By definition each channel length is treated as a single HSU. Two ways are
provided for dividing the hillslope up into HSUS.
The first way is *splitting* where one or more landscape properties are
divided up into classes and HSUs defined by each unique combination of
classes.
The second *burning* enforces classes onto distinct areas.

The `classify` method of a `dynatopGIS` allows the use of both methods, first
*splitting* then *burning* to create hillslope HSUs.

### Splitting

To split a catchment into HRUs the breaks between the classes need to be
specified. This is done by forming a named ```list``` of cuts. The names
correspond to layers of GIS data within the `dynatopGIS` object. The values of
the variables should be numeric and are used as follows:

- `NA`: use the layer 'as is'; i.e. presume it is already classified
- Single value: defines the number of splits which are automatically selected
- Vector of values: treated as the breaks between classes

To demonstate this HSUs for Swindale based on dividing the topographic index
into 21 classes and adding a single class for all areas higher then 500m can
be generated by first creating an additional layer for areas greater then 500 m

```{r height layer}
tmp <- raster::reclassify( ctch$get_layer("filled_dem"), #RasterLayer
                          matrix(c(0,500,NA,
                                   500,1000,-999),
                                   byrow=TRUE))
ctch$add_layer(tmp,'greater_500') ## add the layer to ctch
```
	
then classifying
	
```{r, atb_500_split}
ctch$classify(cuts=list(atanb=20),burns="greater_500")
```

Note that by giving the area to be burnt in negative value we have ensured
that the values do not clash whith those generated by the cuts which will
(except potentially when a cut is NA) will always be positive.

The a list of the resulting parital and final classifications available is
available with

```{r get_class_all}
ctch$get_class()
```

and can be accessed and plotted using `ctch$get_class` and `ctch$plot_layer`
which work in a similar manor `ctch$get_layer` and `ctch$plot_layer`. The
final classificationcan be plotted with 

```{r plot_class}
ctch$plot_class()
```

# Generating a dynamic TOPMODEL

A Dynamic TOPMODEL sutiable for use with the ```dynatop``` package can be
generated using the `create_model` method. This used the current
classification to generate the model. The required model structure is
given in the vignettes of `dynatop` package and is not described here in
details. 

Since `dynatop` simulations make use of ordered HSUs to work downslope, a
seperation by the `band` property which represents the ordered downslope
sequencing of the pixels is used to enforce ordering. The `create_model`
method takes an input in the same format as the specification of a `split`
in the `classify` method. This is applied to the `band` layer as a final stage
in the HSU clasification. Further can be found in the seperate vignette
(TODO).

For example in the case of the division of Swindale by topographic index into
21 classes and the bands into 6 classes the resulting model can be generated by

```{r, model_atb_split}
ctch$classify(cuts=list(atanb=20))
ctch$create_model(band=5)
```

The `get_model` method returns the resulting model in a format suitable for
teh `dynatop` package. The model can be plotted with

```{r plot_model}
ctch$plot_model()
```
