---
title: "Filling sinks and gaps in the DEM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Filling sinks and gaps in the DEM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
                    collapse = TRUE,
                    comment = "#>"
                  )
```

```{r echo=FALSE}
## devtools::load_all("/home/paul/Documents/Software/dynamic_topmodel/dynatopGIS")
library("dynatopGIS")
library("raster")
```
The purpose of this vignette is to outline the sink filling routine used by
the `sink_fill` function and provide some simple demonstrations.

The routine used is deliberately simple, doing the minimum required to populate
sinks and missing data to produce valid flow directions. More complex routines
can by found in many GIS packages and published work.

# Algorithm
In the following we take $z_{k}$ to be the height of the dem at cell $k$, $\mathcal{N}_{k}$ to be the set of index values for the neighbouring
points in space with non-missing
(i.e not `NA` or `NaN` in R)
height values and $\hat{\mathcal{N}}_{k}$ teh set of index values for
neighbouring points with finite heights.

## Identifying data requiring replacement

### Determining missing data in the DEM

To determine which data are missing non-finite values not associated witht he
edge of the DEM are identified. This is done by

1. Replacing all non-finite values in the dem with `NA`
2. Creating clumps (see `raster::clump`) of all `NA` cell values
3. Taking all clumps that do not touch the extent of the raster DEM to be
   missing data.

Cells with missing data are assigned a value of `Inf` ($\infty$). Initialise
$\mathcal{R}$ at set of locations where the DEM value needs replacement with
there location (the associated values of $k$).

### Determining sinks within the DEM

A sink is any pixel in the DEM for which $z_{k} \leq \min_{j \in
\mathcal{N}_k} \left(z_{j}\right)$. A pass through the DEM can be used to
identify sinks. Where a sink is identified set $z_{k}$ to `Inf` ($\infty$), add $k$ to $\mathcal{R}$ and check the neighbouring cells to ensure they are not now sinks.

## Replacing values

Replacing the unknown values from the 'lowest' up would ensure all could be
relaced by a single pass through the points in $\mathcal{R}$. However since
the values are not known this not possible. Instead we approximate this by:

1. For each $k \in \mathcal{R}$ compute $m_k = \min_{j \in
\hat{\mathcal{N}}_k} \left(z_{j}\zright)$
2. Take $\tilde{k}$ and value of $k$ for which $m_k$ is at a minimum.
3. Set $z_k = \mean_{j \in \hat{\mathcal{N}}_k} \left(z_{j}\right)$
4. Remove $k$ from $\mathcal{R}$ and return to step 1

Note that the replacement rule in step 3 ensure that no new sinks are
generated.

. New dem values are computed as the mean of the adjoining finite cell values. New minimums for the adjacent cells are then computed before resorting the sinks and moving on.

# Examples

Example dates is generated randomly

```{r, data}
M <- matrix(1:100,10,10)
fake <- raster::raster(M,
                       crs="+init=epsg:27700") # UK OS projection
```

## handling of a larger area of missing

```{r, example1}
ex1_dem <- fake
ex1_dem[c(23:26,33:36,44:45)] <- NA
ex1_chn <- fake; ex1_chn[] <- NA
ex1_chn[1,1] <- 1


ex1 <- dynatopGIS$new(ex1_dem)
ex1$add_layer(ex1_chn,"channel_id",FALSE)
ex1$sink_fill()
ex1$plot_layer("dem",FALSE)
ex1$plot_layer("filled_dem",FALSE)
```

## handling of a single isleland

```{r, example2}
ex2_dem <- fake
ex2_dem[c(23:26,33:36,44:45)] <- NA
ex2_chn <- fake; ex2_chn[] <- NA
ex2_chn[1,1] <- 1

ex2 <-  dynatopGIS$new(ex1_dem)
ex2$add_layer(ex2_chn,"channel_id",FALSE)
ex2$sink_fill()
ex2$plot_layer("dem",add=FALSE)
ex2$plot_layer("filled_dem",add=FALSE)
```
